HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-24 scnn.tex
added figure

\begin{figure*}
	\centering
    % \includegraphics[width=\linewidth]{architecture}
	\includegraphics[width=\linewidth]{figure_architecture}
	\caption{Overall network architecture, showing here two convolutional layers acting as feature extractors followed by a fully connected layer with softmax acting as the classifier. \nati{add the stat layer?} A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling. While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted. In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph. Graphs are versatile data structures which can model any sampling, even irregular or partial. Second, we propose to exploit a hierarchical pixelization of the sphere for the pooling operation. It allows the network to analyze the data at multiple scales while preserving the spatial localization of features. This figure shows a network which operates on the whole sphere. The process is however the same when working with parts of sphere, except that the graph is only built for the part of interest.}
	\label{fig:architecture}
\end{figure*}

@@

To overcome those shortcomings, we introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling. The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels. The length (or weight) of the shortest path between two pixels is an approximation of the geodesic distance between them on the manifold.
This contribution leverages the emerging field of graph signal processing which defines fundamental signal processing tools such as the Fourier transform for signals defined on graphs \citep{shuman2013emerging}.
Our approach uses the graph CNN formulation introduced by \citet{defferrard2016convolutional}, which is efficient and can adapt to any sampling of the sphere. Using a $k$-nearest neighbors graph, the convolution operation costs $\mathcal{O}(N_{pix})$ operations, where $N_{pix}$ is the number of pixels. This is the lowest possible complexity for a convolution operation without approximations (e.g. by sketching).
The flexibility of modeling the data domain with a graph allows, for example, to easily model data which only spans a part of the sphere, or data which is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelization of the manifold to analyze the data at multiple scales.
Such hierarchical pixelizations of the sphere have been well studied.

@@

Following \citet{defferrard2016convolutional}, we define our filters as Chebyshev polynomials such as the convolution operation becomes
\begin{equation} \label{eqn:graph_convolution_cheby}
	h_\theta(\tL) \f = \sum_{k=0}^K \theta_k T_k(\tL) \f,
\end{equation}
where $\tL = \frac{2}{\lambda_{max}} \L - \b{I} = -\frac{2}{\lambda_{max}} \D^{-1/2} \W \D^{-1/2}$ is the rescaled Laplacian with eigenvalues in $[-1, 1]$ and $T_k(\cdot)$ is the Chebyshev polynomial of degree $k$, defined by the recursive relation $T_k(\tL) = 2\tL T_{k-1}(\tL) - T_{k-2}(\tL)$, $T_1(\tL) = \tL$, $T_0(\tL) = \b{I}$. While definitions \eqnref{graph_convolution_polynomial} and \eqnref{graph_convolution_cheby} both allow the representation of the same set of filters, Chebyshev polynomials are preferred as they form an orthogonal basis on the interval $[-1, 1]$. Moreover, they factor out $\tL^{k-2}$ from $T_k(\tL)$, e.g., the 3-neighborhood is not mixed with the 5-neighborhood. That results in easier learning when optimizing the parameters $\theta$.
Finally, the coefficients can be computed such as the truncated Chebyshev expansions gives an approximate minimax polynomial of a given filter \citep{hammond2011wavelets}. While that is irrelevant when learning filters, it is useful when smoothing a signal on the sphere by filtering it with an approximated low-pass filter.
\todo{(i) scale of $\theta_k$: they are mostly uniform for Chebyshev polynomials, end exploding for monomials. Though SGD implicitly imposes a uniform prior on the learned parameters.}
\todo{(ii) numerical stability of $\L^k \f$ v.s. $T_k(\L) \f$}
\todo{(iii) in the code, initialize $\theta_k$ with stdev adapted to $F_{in}$ and $K$}

@@

Given a map $\X \in \R^{N_{pix} \times F}$, the neural network is computing $\y = NN_\theta(\X) \in R^{N_{classes}}$, where $NN$ is a composition of the above layers (an example is $NN = SM \circ FC \circ \sigma \circ FC \circ ST \circ \sigma \circ GC \circ P \circ \sigma \circ BN \circ GC$) and $\theta$ is the set of all trainable parameters.
The number of features $F$ depends on the data at hand. For the CMB radiation temperature, $F = 1$. For observations in radio frequencies, $F$ would be equal to the number of surveyed frequencies. $F$ might also be the number of slices in the radial direction.
The loss function $\ell(\y) = \ell(NN_\theta(\X))$ is a measure of how good the prediction $\y$ is for the sample $\X$. For a classification task, the loss is usually taken to be the cross-entropy
\begin{equation*}
	\ell(\y) = -\log(y_i),
\end{equation*}
where $i$ is the correct class label.
The parameters $\theta$ of the neural network are optimized to minimize the loss over the training dataset $\left\{ \X_i \right\}_{i=1}^{N_{samples}}$:
\begin{equation*}
	\hat{\theta} = \argmin_\theta \sum_{i=1}^{N_{samples}} \ell(NN_\theta(\X_i)).
\end{equation*}
The optimization is performed by computing an error gradient w.r.t.\ all the parameters by back-propagation and updating them with a form of stochastic gradient descent (SGD):
\begin{equation*}
	\theta \leftarrow \theta - \frac{\eta}{|\B|} \sum_{i \in \B} \frac{\partial \ell(NN_\theta(\X_i))}{\partial \theta} ,
\end{equation*}
where $\eta$ is the learning rate, and $\B$ is the set of indices in a mini-batch. Batches are used instead of single samples to gain speed by exploiting the parallelism afforded by modern computing platforms.

@@

% Structure of the section
% 1) Present the goal
% 2) Explain briefly the dataset
% 3) Present the setting
% 4) Describe the result
In this section, we demonstrate the performance of the spherical neural network on a discrimination problem: the classification of convergence maps, similar to those created with gravitational lensing technique \citep{chang2017curvedsky}. Nevertheless, by simply changing the loss, one could use the network for regression.
Both our Python implementation\footnote{\url{https://github.com/SwissDataScienceCenter/HealpixNet}} and the data\footnote{\url{https://doi.org/10.5281/zenodo.1303272}} to reproduce those experiments is available online.

>>>>

\begin{figure*}
	\centering
    % \includegraphics[width=\linewidth]{architecture}
	\includegraphics[width=\linewidth]{figure_architecture_v2}
	\caption{Overall network architecture, showing here two convolutional layers acting as feature extractors followed by a fully connected layer with softmax acting as the classifier. \nati{add the stat layer?} A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling. While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted. In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph. Graphs are versatile data structures which can model any sampling, even irregular or partial. Second, we propose to exploit a hierarchical pixelization of the sphere for the pooling operation. It allows the network to analyze the data at multiple scales while preserving the spatial localization of features. This figure shows a network which operates on the whole sphere. The process is however the same when working with parts of sphere, except that the graph is only built for the part of interest.}
	\label{fig:architecture}
\end{figure*}

@@

To overcome those shortcomings, we introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling. The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels.
This contribution leverages the emerging field of graph signal processing which defines fundamental signal processing tools such as the Fourier transform for signals defined on graphs \citep{shuman2013emerging}.
Our approach uses the graph CNN formulation introduced by \citet{defferrard2016convolutional}, which is efficient and can adapt to any sampling of the sphere. Using a $k$-nearest neighbors graph, the convolution operation costs $\mathcal{O}(N_{pix})$ operations, where $N_{pix}$ is the number of pixels. This is the lowest possible complexity for a convolution operation without approximations (e.g. by sketching).
The flexibility of modeling the data domain with a graph allows, for example, to easily model data which only spans a part of the sphere, or data which is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelization of the manifold to analyze the data at multiple scales.
Such hierarchical pixelizations of the sphere have been well studied.

@@

Following \citet{defferrard2016convolutional}, we define our filters as Chebyshev polynomials such as the convolution operation becomes
\begin{equation} \label{eqn:graph_convolution_cheby}
	h_\theta(\tL) \f = \sum_{k=0}^K \theta_k T_k(\tL) \f,
\end{equation}
where $\tL = \frac{2}{\lambda_{max}} \L - \b{I} = -\frac{2}{\lambda_{max}} \D^{-1/2} \W \D^{-1/2}$ is the rescaled Laplacian with eigenvalues in $[-1, 1]$ and $T_k(\cdot)$ is the Chebyshev polynomial of degree $k$, defined by the recursive relation $T_k(\tL) = 2\tL T_{k-1}(\tL) - T_{k-2}(\tL)$, $T_1(\tL) = \tL$, $T_0(\tL) = \b{I}$. While definitions \eqnref{graph_convolution_polynomial} and \eqnref{graph_convolution_cheby} both allow the representation of the same set of filters, Chebyshev polynomials are preferred as they form an orthogonal basis on the interval $[-1, 1]$. Moreover, they factor out $\tL^{k-2}$ from $T_k(\tL)$, e.g., the 3-neighborhood is not mixed with the 5-neighborhood. That results in easier learning when optimizing the parameters $\theta$.
Finally, the coefficients can be computed such as the truncated Chebyshev expansions gives an approximate minimax polynomial of a given filter \citep{hammond2011wavelets}. While that is irrelevant when learning filters, it is useful when smoothing a signal on the sphere by filtering it with an approximated low-pass filter.
\todo{(i) scale of $\theta_k$: they are mostly constant for Chebyshev polynomials. How are they for monomials?}
\todo{(ii) numerical stability of $\L^k \f$ v.s. $T_k(\L) \f$}
\todo{(iii) in the code, initialize $\theta_k$ with stdev adapted to $F_{in}$ and $K$}

@@

Given a map $\X \in \R^{N_{pix} \times F}$, the neural network is computing $\y = NN_\theta(\X) \in R^{N_{classes}}$, where $NN$ is a composition of the above layers (an example is $NN = SM \circ FC \circ \sigma \circ FC \circ ST \circ \sigma \circ GC \circ P \circ \sigma \circ BN \circ GC$) and $\theta$ is the set of all trainable parameters.
The number of features $F$ depends on the data at hand. For the CMB radiation temperature, $F = 1$. For observations in radio frequencies, $F$ would be equal to the number of surveyed frequencies. $F$ might also be the number of slices in the radial direction.
The loss function $\ell(\y) = \ell(NN_\theta(\X))$ is a measure of how good the prediction $\y$ is for the sample $\X$. For a classification task, the loss is usually taken to be the cross-entropy
\begin{equation*}
	\ell(\y) = -\log(y_i),
\end{equation*}
where $i$ is the correct class label.
The parameters $\theta$ of the neural network are optimized to minimize the loss over the training dataset $\left\{ \X_i \right\}_{i=1}^{N_{samples}}$:
\begin{equation*}
	\hat{\theta} = \argmin_\theta \sum_{i=1}^{N_{samples}} \ell(NN_\theta(\X_i)).
\end{equation*}
The optimization is performed by computing an error gradient w.r.t.\ all the parameters by back-propagation and updating them with a form of stochastic gradient descent (SGD):
\begin{equation*}
	\theta^{(k+1)} = \theta^{(k)} - \frac{\eta}{|\B|} \sum_{i \in \B} \frac{\partial \ell(NN_\theta(\X_i))}{\partial \theta} ,
\end{equation*}
where $\eta$ is the learning rate, and $\B$ is the set of indices in a mini-batch. Batches are used instead of single samples for faster training.

@@

% Structure of the section
% 1) Present the goal
% 2) Explain briefly the dataset
% 3) Present the setting
% 4) Describe the result
In this section, we demonstrate the performance of the spherical neural network on a discrimination problem. Nevertheless, by simply changing the loss, one could use the network for regression.
Our implementation is available as a small python package \url{https://our.code.link.com}.
Furthermore, all the data to reproduce our experiments are available online.\footnote{\url{https://doi.org/10.5281/zenodo.1303272} \todo{Check DOI}}.

In our experiments we classify convergence maps, similar to those created with gravitational lensing technique \citep{chang2017curvedsky}.
